Sentence Generation as Planning with Probabilistic LTAG\\

### 2010-05-20 trunk/Papers/pcrisp-10/crisp.tex
Spell checked

\section{Sentence Generation as Planning}
\label{sec:crisp}
In this section we review the original non-statistical {\sc crisp} system \cite{kollerstone2007}. 
Following {\sc spud} \cite{stonedoran1997}, {\sc crisp} is based on a declarative description of the sentence generation problem using {\sc tag}. Given a knowledge base, a communicative goal and a grammar, we require to find a grammatical {\sc tag} derivation that is consistent with this knowledge base and satisfies a communicative goal. A number of semantic and pragmatic constraints that must be satisfied by the solution can be added, for instance to enforce generation of unambiguous refering expressions. \newcite{kollerstone2007} describe how to encode this problem into an AI planning problem, that can be solved efficiently by off-the-shelf planners. We describe the general mechanism in the following section and then review the encoding into planning in section \ref{ssec:crispdomain}. 
\begin{figure*}[th]
\begin{center}
\includegraphics[width=.8\textwidth]{figures/grammar.pdf}
\caption{\label{fig:grammar}(a) An example grammar with semantic content. (b) A derivation for ``the cat eats the raw fish''. }
\end{center}
\end{figure*}

@@

In a derivation we may only include elementary trees whose semantic content has an instantiation in the knowledge base. For each substitution and adjunction, the semantic role associated with the role of the target node is unified with the `self' role of the child tree. For example, given the knowledge base 
{\it \{cat($i_2$), fish($i_3$), raw($i_3$), fish($i_4$), eats($i_1$,$i_2$,$i_3$)\}} and the grammar in \ref{fig:grammar}(a), {\sc crisp} could produce the derivation in \ref{fig:grammar}(b). Notice that {\sc crisp} can generate the unambiguos refering expression `the raw fish' for $i_3$, to distinguish it from $i_4$. For details on  refering expression generation in {\sc crisp} see \newcite{garoufikoller2010}.

@@

Figure \ref{fig:crisp-operators} shows planning operators for part of the grammar in figure \ref{fig:grammar}(a).
The preconditions of the operators require that a suitable open substitution node (i.e. of the correct category) or internal node for adjunction exists in the partial derivation. In the operator effect, open substitution nodes are closed and new identifiers are created for each substitution node and internal node in the new tree. Given the knowlede base from above, a plan corresponding to the derivation in figure \ref{fig:grammar} would be $\langle${\it S-eats-1(root, i1, i2, i3), NP-cat-2(subj-1, i2), NP-fish-3(obj-1, i3), N-raw-4(obj-1, i3)}$\rangle$

\subsection{CRISP and Large Grammars}
Using {\sc crisp} with large, treebank induced grammars can lead to overgeneration, 
%Such trees contain many internal nodes that are open for adjunction but will not necessarily be filled in a derivation for a given input communicative goal. Attaching to the wrong node can produce invalid or disprefered word order. 
due to the high lexical ambiguity of possible, structurally different trees (supertags) for a given word. We illustrate this problem in figure \ref{fig:overgen}. Assuming a grammar that includes trees for both right adjoining ($t13$) and left adjoining PPs ($t252$), both derivations (a) and (b) are {\it grammatical} derivations that satisfy the same communicative goal. However, most readers disprefer the reading in (b). 
Clearly, to use {\sc crisp} with such a grammar we need a method of distinguishing good derivations from bad ones. In section \ref{sec:experiments} we provide empirical evidence for this problem and show how it is solved by the approach we describe in the next section. 
\begin{figure}
\begin{center}
\includegraphics[width=.3\textwidth]{figures/overgen.pdf}
\caption{\label{fig:overgen} Two derivations with a large grammar, that satisfy the same communicative goal. Sentence (b) is disprefered by most readers.}
\end{center}
\end{figure}

>>>>

\section{Sentence Generation as Planning}
\label{sec:crisp}
In this section we review the original non-statistical {\sc crisp} system \cite{kollerstone2007}. 
Following {\sc spud} \cite{stonedoran1997}, {\sc crisp} is based on a declarative description of the sentence generation problem using {\sc tag}. Given a knowledge base, a communicative goal and a grammar, we require to find a grammatical {\sc tag} derivation that is consistent with this knowledge base and satisfies a communicative goal. A number of semantic and pragmatic constraints that must be satisfied by the solution can be added, for instance to enforce generation of unambiguous referring expressions. \newcite{kollerstone2007} describe how to encode this problem into an AI planning problem, that can be solved efficiently by off-the-shelf planners. We describe the general mechanism in the following section and then review the encoding into planning in section \ref{ssec:crispdomain}. 
\begin{figure*}[th]
\begin{center}
\includegraphics[width=.8\textwidth]{figures/grammar.pdf}
\caption{\label{fig:grammar}(a) An example grammar with semantic content. (b) A derivation for ``the cat eats the raw fish''. }
\end{center}
\end{figure*}

@@

In a derivation we may only include elementary trees whose semantic content has an instantiation in the knowledge base. For each substitution and adjunction, the semantic role associated with the role of the target node is unified with the `self' role of the child tree. For example, given the knowledge base 
{\it \{cat($i_2$), fish($i_3$), raw($i_3$), fish($i_4$), eats($i_1$,$i_2$,$i_3$)\}} and the grammar in \ref{fig:grammar}(a), {\sc crisp} could produce the derivation in \ref{fig:grammar}(b). Notice that {\sc crisp} can generate the unambiguous referring expression `the raw fish' for $i_3$, to distinguish it from $i_4$. For details on  referring expression generation in {\sc crisp} see \newcite{garoufikoller2010}.

@@

Figure \ref{fig:crisp-operators} shows planning operators for part of the grammar in figure \ref{fig:grammar}(a).
The preconditions of the operators require that a suitable open substitution node (i.e. of the correct category) or internal node for adjunction exists in the partial derivation. In the operator effect, open substitution nodes are closed and new identifiers are created for each substitution node and internal node in the new tree. Given the knowledge base from above, a plan corresponding to the derivation in figure \ref{fig:grammar} would be $\langle${\it S-eats-1(root, i1, i2, i3), NP-cat-2(subj-1, i2), NP-fish-3(obj-1, i3), N-raw-4(obj-1, i3)}$\rangle$

\subsection{CRISP and Large Grammars}
Using {\sc crisp} with large, treebank induced grammars can lead to overgeneration, 
%Such trees contain many internal nodes that are open for adjunction but will not necessarily be filled in a derivation for a given input communicative goal. Attaching to the wrong node can produce invalid or dispreferred word order. 
due to the high lexical ambiguity of possible, structurally different trees (supertags) for a given word. We illustrate this problem in figure \ref{fig:overgen}. Assuming a grammar that includes trees for both right adjoining ($t13$) and left adjoining PPs ($t252$), both derivations (a) and (b) are {\it grammatical} derivations that satisfy the same communicative goal. However, most readers disprefer the reading in (b). 
Clearly, to use {\sc crisp} with such a grammar we need a method of distinguishing good derivations from bad ones. In section \ref{sec:experiments} we provide empirical evidence for this problem and show how it is solved by the approach we describe in the next section. 
\begin{figure}
\begin{center}
\includegraphics[width=.3\textwidth]{figures/overgen.pdf}
\caption{\label{fig:overgen} Two derivations with a large grammar, that satisfy the same communicative goal. Sentence (b) is dispreferred by most readers.}
\end{center}
\end{figure}

### 2010-05-20 trunk/Papers/pcrisp-10/experiments.tex
Spell checked

\section{Experiments}
\label{sec:experiments}
In this section we perform experiments with the statistical generation system described in the previous section.  We are interested in the output quality and runtime behavior of different combinations of planning systems and probability models. Like {\sc crisp}, {\sc pcrisp} is a sentence generation system and can for instance generate refering expressions. Unfortunately, since we do not have an evaluation dataset for full sentence generation, we only evaluate the realization abilities of our system by comparing generated sentences against a gold standard testset. The next section describes our evaluation data and section \ref{ssec:eval-measures} describes the evaluation measures used.

@@

\subsection{Evaluation Measures}
\label{ssec:eval-measures}
To evaluate the ouput quality of our statistical generation system we compare the system output $O$ against the reference sentence $R$in the treebank, from which the input representation was generated. We adopt the {\bf generation tree accuracy} measure discussed by \cite{bangalorerambowwhittaker2001}.
This measure is computed by first creating a list of all 'treelets' from the reference derivation tree $D$. A 'treelet' is a subtree, consisting only of a node and its direct children. For each treelet we calculate edit distance, sum the distances over all treelets and then divide by the total length of the reference string.
$$1 - \frac{\sum\limits_{d \in treelets(D)} editDist(O|_d, R|_d)}{\textit{length}(R)},$$
where $D$ is the reference derivation tree and $S|_d$ are the tokens associated with the nodes of treelet $d$ in the order they appear in $S$ (if at all). 
Edit distance is modified to treat insertion-deletion pairs as single movement errors. 
This metric account for the observation that errors that change word order on a local level (i.e. within a treelet) are less severe than errors that swap words between different treelets. Intuitively errors of the latter type violate the continuity of sub-phrases. Because of this property we prefer generation tree accuracy over {\sc blue}, which is overly sensitive to movement errors on an individual sentence level.

%\newcite{belzreiter2006} empirically investigate the usefulness of {\sc bleu} and {\sc nist} for evaluating NLG systems. They find that both metrics correlate with human judgements of output quality on a corpus level, but require a large number of different references sentences to be reliable. Multiple references are necessary because {\sc blue} favors sentences with the same word order as the reference. In NLG, more than in Machine Translation, word order of output sentences can vary.

@@

%Tree accuracy measure only the quality of the surface string. To compare the syntactic structure of the output with the reference we propose {\it derivation tree accuracy}, the ratio of derivation-treelets (nodes with their direct children in their specific order) that occurs in both the system output derivation and the reference derivation. 
%$$ \frac{|treelets(D_O) \cap treelets(D_R)|}{\textit{length}(D_R)}$$
% Notice that measures used in statistical parsing, e.g. bracketing and labelling accuracy, cannot be used to evaluate generation output because they can only measure similarity between two different syntactic structures for the same surface string.

@@

The first two columns confirm our hypothesis that {\sc crisp} does not scale up to the automatically induced treebank. The greedy {\sc spud} heuristic does not generate a single correct sentence, as the combinatorics of the planning domain becomes too large with the treebank induced grammar. 
With the standard {\sc pltag} definition our system manages to generate only very few sentences due to data sparseness. However these sentences are almost identical to the ones found in the test set. Both generalized models ease the data sparseness problem, but output quality drops rapidly. For the unlexicalized model the search problem becomes very hard due to the large planning domain, so that 22\% of the generation problems require more than 5 minutes runtime. As expected the smoothed model offers a compromise between efficiency, output quality and coverage 
\footnote{In our experiments we set the weight between level 1 and 2 to 0.9 and did not consider level 3, after empirically testing different settings. We set a probability threshold to include operators into the planning domain to $10^{-4}$}.

\begin{table}
    \caption{\label{results} Evaluation results.}
    \begin{center}
    \begin{tabular}{|l|l|r|r|r|}
    \multicolumn{5}{c}{\sc crisp}\\ 
    \hline
     {\bf Planner }  & {\bf acc}& {\bf success} & {\bf fail} & {\bf timeout} \\ \hline 
     greedy         & ??? & 0\% & 100\% & 0\%  \\ \hline 
     FF         &  0.66 & 88\% & 0\% & 12\% \\ \hline 
    \multicolumn{5}{c}{\bf planning with {\sc pltag}}\\\hline
    {\bf Planner }  & {\bf acc}& {\bf success} & {\bf fail} & {\bf timeout} \\ \hline 
     FF + costs &  0.99 &  97\% & 0\% & 3\% \\ \hline
     FF         &  0.99 & 71\% & 0\% &  29\% \\ \hline
    \multicolumn{5}{c}{\bf planning with {\sc pltag} (unlexicalized)}\\\hline
    {\bf Planner }  & {\bf acc}& {\bf success} & {\bf fail} & {\bf timeout} \\ \hline 
     FF + costs & 0.74 & 62\% & 16\% & 22\%\\ \hline
     FF & 0.76 & 62\% & 16\% &  22\%\\ \hline
    \multicolumn{5}{c}{\bf planning with {\sc pltag} (smoothed)}\\\hline
    {\bf Planner }  & {\bf acc}& {\bf success} & {\bf fail} & {\bf timeout} \\ \hline 
     FF + costs & 0.78 & 19\% & 74\% & 7\% \\ \hline
     FF & 0.77 & 19\% & 74\% & 7\% \\ \hline
    \end{tabular}
    \end{center}
\end{table}

>>>>

\section{Experiments}
\label{sec:experiments}
In this section we perform experiments with the statistical generation system described in the previous section.  We are interested in the output quality and runtime behavior of different combinations of planning systems and probability models. Like {\sc crisp}, {\sc pcrisp} is a sentence generation system and can for instance generate referring expressions. Unfortunately, since we do not have an evaluation dataset for full sentence generation, we only evaluate the realization abilities of our system by comparing generated sentences against a gold standard testset. The next section describes our evaluation data and section \ref{ssec:eval-measures} describes the evaluation measures used.

@@

\subsection{Evaluation Measures}
\label{ssec:eval-measures}
To evaluate the output quality of our statistical generation system we compare the system output $O$ against the reference sentence $R$in the treebank, from which the input representation was generated. We adopt the {\bf generation tree accuracy} measure discussed by \cite{bangalorerambowwhittaker2001}.
This measure is computed by first creating a list of all 'treelets' from the reference derivation tree $D$. A 'treelet' is a subtree, consisting only of a node and its direct children. For each treelet we calculate edit distance, sum the distances over all treelets and then divide by the total length of the reference string.
$$1 - \frac{\sum\limits_{d \in treelets(D)} editDist(O|_d, R|_d)}{\textit{length}(R)},$$
where $D$ is the reference derivation tree and $S|_d$ are the tokens associated with the nodes of treelet $d$ in the order they appear in $S$ (if at all). 
Edit distance is modified to treat insertion-deletion pairs as single movement errors. 
This metric account for the observation that errors that change word order on a local level (i.e. within a treelet) are less severe than errors that swap words between different treelets. Intuitively errors of the latter type violate the continuity of sub-phrases. Because of this property we prefer generation tree accuracy over {\sc blue}, which is overly sensitive to movement errors on an individual sentence level.

%\newcite{belzreiter2006} empirically investigate the usefulness of {\sc bleu} and {\sc nist} for evaluating NLG systems. They find that both metrics correlate with human judgments of output quality on a corpus level, but require a large number of different references sentences to be reliable. Multiple references are necessary because {\sc blue} favors sentences with the same word order as the reference. In NLG, more than in Machine Translation, word order of output sentences can vary.

@@

%Tree accuracy measure only the quality of the surface string. To compare the syntactic structure of the output with the reference we propose {\it derivation tree accuracy}, the ratio of derivation-treelets (nodes with their direct children in their specific order) that occurs in both the system output derivation and the reference derivation. 
%$$ \frac{|treelets(D_O) \cap treelets(D_R)|}{\textit{length}(D_R)}$$
% Notice that measures used in statistical parsing, e.g. bracketing and labeling accuracy, cannot be used to evaluate generation output because they can only measure similarity between two different syntactic structures for the same surface string.

@@

The first two columns confirm our hypothesis that {\sc crisp} does not scale up to the automatically induced treebank. The greedy {\sc spud} heuristic does not generate a single correct sentence, as the combinatorics of the planning domain becomes too large with the treebank induced grammar. 
With the standard {\sc ptag} definition our system manages to generate only very few sentences due to data sparseness. However these sentences are almost identical to the ones found in the test set. Both generalized models ease the data sparseness problem, but output quality drops rapidly. For the unlexicalized model the search problem becomes very hard due to the large planning domain, so that 22\% of the generation problems require more than 5 minutes runtime. As expected the smoothed model offers a compromise between efficiency, output quality and coverage 
\footnote{In our experiments we set the weight between level 1 and 2 to 0.9 and did not consider level 3, after empirically testing different settings. We set a probability threshold to include operators into the planning domain to $10^{-4}$}.

\begin{table}
    \caption{\label{results} Evaluation results.}
    \begin{center}
    \begin{tabular}{|l|l|r|r|r|}
    \multicolumn{5}{c}{\sc crisp}\\ 
    \hline
     {\bf Planner }  & {\bf acc}& {\bf success} & {\bf fail} & {\bf timeout} \\ \hline 
     greedy         & ??? & 0\% & 100\% & 0\%  \\ \hline 
     FF         &  0.66 & 88\% & 0\% & 12\% \\ \hline 
    \multicolumn{5}{c}{\bf planning with {\sc ptag}}\\\hline
    {\bf Planner }  & {\bf acc}& {\bf success} & {\bf fail} & {\bf timeout} \\ \hline 
     FF + costs &  0.99 &  97\% & 0\% & 3\% \\ \hline
     FF         &  0.99 & 71\% & 0\% &  29\% \\ \hline
    \multicolumn{5}{c}{\bf planning with {\sc ptag} (unlexicalized)}\\\hline
    {\bf Planner }  & {\bf acc}& {\bf success} & {\bf fail} & {\bf timeout} \\ \hline 
     FF + costs & 0.74 & 62\% & 16\% & 22\%\\ \hline
     FF & 0.76 & 62\% & 16\% &  22\%\\ \hline
    \multicolumn{5}{c}{\bf planning with {\sc ptag} (smoothed)}\\\hline
    {\bf Planner }  & {\bf acc}& {\bf success} & {\bf fail} & {\bf timeout} \\ \hline 
     FF + costs & 0.78 & 19\% & 74\% & 7\% \\ \hline
     FF & 0.77 & 19\% & 74\% & 7\% \\ \hline
    \end{tabular}
    \end{center}
\end{table}

### 2010-05-20 trunk/Papers/pcrisp-10/pcrisp.tex
Related Work
Spell checked

We now extend {\sc crisp} to statistical generation ({\sc pcrisp}). The basic idea is to add a statistical grammar model while leaving the sentence generation mechanism untouched. This way we can select the highest scoring derivation, that satisfies all constraints (grammaticality, expresses the communicative goal, use unambiguous refering expressions etc.).

 As a straightforward probability model over {\sc ltag} derivations we choose probabilistic {\sc tag} ({\sc ptag}) \cite{resnik1992}.
Our choice of {\sc ptag} for sentence generation is motivated by a number of attractive properties.
 {\sc Ptag} is lexicalized and therefore does not only assign probabilities to operations in the grammar (as for example plain {\sc pcfg}), but also accounts for binary dependencies between words.  Unlike n-gram models however, these co-occurences are structured according to local syntactic context as a result of {\sc tag}'s extended domain of locality. The probability model describes how the syntactic arguments of a word are typically filled. 
Furthermore, as {\sc tag} factors recursion from the domain of dependencies, the probability for core constructions remains the same independent of additional adjunctions. 
%Second, {\sc ptag} is a generative model. This is crucial because it allows us to assign probabilities to partial derivations. In addition we use the probability of partial derivations as a cost function, to help guide search through the space of possible PLTAG derivations. In contrast, other statistical generation systems use discriminative models that select the best possible generation output from a set of candidate sentences.
%Finally, the independence assumption makes it easy to estimate {\sc ptag}s from a treebank.
 We review {\sc ptag} in section \ref{ssec:probmodels}.

@@

$$ P_{adj}(\adj(\tau, lex_\tau, \textit{None},n)) +$$
$$ \sum\limits_{\beta \in A}\sum\limits_{w_\beta} P_{adj}(\adj(\tau, w, \tau, v, n)) = 1.$$

@@

\subsection{PCRISP Planning Domains}
\label{ssec:pcrisp-domains}
Using the definition of {\sc ptag}, we now reformulate the {\sc crisp} planning operators described in section \ref{ssec:crispdomain}.
The independence assumption in {\sc ptag} allows us to maintain that planning operator add a single tree, but are now assigned a probability score. However, while {\sc crisp} planning operators can add an elementary tree to any site of the right category, {\sc ptag} substitution and adjunction events are binary events between lexicalized trees at a specific node. We therefore adapt the literals that record open substitution and adjunction sites in partial derivations accordingly and create one operator for each node in each possible combination of lexicalized trees.
Figure \ref{example-action} shows an example planning operator for each type.
\begin{figure}[t]
\begin{center}
\cplanaction{\bf subst-t3-cat-t28-eats-n1(u,~x1)}{referent(u,~x1),\\ subst(t28-eats,~n1,~u), cat(x1)}{
$\lnot$needtoexpress(pred-cat,~x1),\\ $\lnot$subst(t-28-eats, n1,~u),\\
adj(t3-cat, n2 u)}{4.3012}\\\smallskip

\cplanaction{\bf adj-t5-raw-t3-fish-n2(u,~x1)}{referent(u,~x1),\\ adj(t28-eats,n2,~u), raw(x1)}{
$\lnot$needtoexpress(pred-raw,~x1),\\ $\lnot$adj(t-28-eats, n2,~u)}{6.9076}\\\smallskip

\cplanaction{\bf init-t28-eats(u,~x1,~x2,~x3)}{referent(u,~x1),\\ eats(x1,~x2,~x3)}{
$\lnot$needtoexpress(pred-eats,~x1,~x2,~x3), \\
 subst(t-28-eats,~n1,subj), subst(t28-eats,~n4,obj),\\
adj(t-28-eats, n2, u), adj(t-28-eats, n3, u)}{8.5172}\\\smallskip

@@

  A straightforward way to deal with data sparseness is to drop all lexicalization from event definitions, as illustrated in figure \ref{fig:modelillustration}, A. Unfortunately this model does no longer account for bilexical dependencies between words. Since our system has to add a lexicalized tree in each step, lexicalization for this child tree should always be taken into account by the probability model, if available. Despite these drawbacks, we perform experiments with the unlexicalized model as a baseline. This allows us to investigate if purely syntactic information is sufficient to achieve high quality generation output. \\  
An alternative model computes linear interpolation between three back-off levels. The first level is just standard {\sc ptag} (figure \ref{fig:modelillustration}, B.1), for the second level the lexicalization of the parent tree is dropped (figure \ref{fig:modelillustration}, B.2), for the third level the model describes only the distribution of lexicalized child trees over each category (figure \ref{fig:modelillustration}, B.3).  
Notice also, that the third level is similar in spirit to a probabilistic version of original {\sc crisp} operators (compare figure \ref{fig:modelillustration}, B.3 and figure \ref{fig:crisp-operators}. 
\begin{figure}[t]
\begin{center}
\includegraphics[width=.5\textwidth]{figures/modelillustration}
\caption{\label{fig:modelillustration} Illustration of the unlexicalized probability model (A) and the three back-off levels of the linear interpolation model (B). B.1 is the original {\sc ptag} definition and corresponds directly to {\sc pcrisp} operators as in figure \ref{example-action}, while B.3 resembles {\sc crisp} operators (c.f. figure \ref{fig:crisp-operators}).}

>>>>

We now extend {\sc crisp} to statistical generation ({\sc pcrisp}). The basic idea is to add a statistical grammar model while leaving the sentence generation mechanism untouched. This way we can select the highest scoring derivation, that satisfies all constraints (grammaticality, expresses the communicative goal, use unambiguous referring expressions etc.).

 As a straightforward probability model over {\sc ltag} derivations we choose probabilistic {\sc tag} ({\sc ptag}) \cite{resnik1992}.
Our choice of {\sc ptag} for sentence generation is motivated by a number of attractive properties.
 {\sc Ptag} is lexicalized and therefore does not only assign probabilities to operations in the grammar (as for example plain {\sc pcfg}), but also accounts for binary dependencies between words.  Unlike n-gram models however, these co-occurrences are structured according to local syntactic context as a result of {\sc tag}'s extended domain of locality. The probability model describes how the syntactic arguments of a word are typically filled. 
Furthermore, as {\sc tag} factors recursion from the domain of dependencies, the probability for core constructions remains the same independent of additional adjunctions. 
%Second, {\sc ptag} is a generative model. This is crucial because it allows us to assign probabilities to partial derivations. In addition we use the probability of partial derivations as a cost function, to help guide search through the space of possible PLTAG derivations. In contrast, other statistical generation systems use discriminative models that select the best possible generation output from a set of candidate sentences.
%Finally, the independence assumption makes it easy to estimate {\sc ptag}s from a treebank.
 We review {\sc ptag} in section \ref{ssec:probmodels}.

@@

$$ P_{adj}(\adj(\tau, v, \textit{None}, n)) +$$
$$ \sum\limits_{\beta \in A}\sum\limits_{w_\beta} P_{adj}(\adj(\tau, v, \beta, w, n)) = 1.$$

@@

\subsection{PCRISP Planning Domains}
\label{ssec:pcrisp-domains}
Using the definition of {\sc ptag}, we now reformulate the {\sc crisp} planning operators described in section \ref{ssec:crispdomain}.
The independence assumption in {\sc ptag} allows us to maintain that planning operator add a single tree, but are now assigned a probability score. However, while {\sc crisp} planning operators can add an elementary tree to any site of the right category, {\sc ptag} substitution and adjunction events are binary events between lexicalized trees at a specific node. We therefore adapt the literals that record open substitution and adjunction sites in partial derivations accordingly and create one operator for each node in each possible combination of lexicalized trees.
Figure \ref{example-action} shows an example planning operator for each type.
\begin{figure}[t]
\begin{center}
\cplanaction{\bf subst-t3-cat-t28-eats-n1(u,~x1)}{referent(u,~x1),\\ subst(t28-eats,~n1,~u), cat(x1)}{
$\lnot$needtoexpr(pred-cat,~x1),\\ $\lnot$subst(t-28-eats, n1,~u),\\
adj(t3-cat, n2 u)}{4.3012}\\\smallskip

\cplanaction{\bf adj-t5-raw-t3-fish-n2(u,~x1)}{referent(u,~x1),\\ adj(t28-eats,n2,~u), raw(x1)}{
$\lnot$needtoexpr(pred-raw,~x1),\\ $\lnot$adj(t-28-eats, n2,~u)}{6.9076}\\\smallskip

\cplanaction{\bf init-t28-eats(u,~x1,~x2,~x3)}{referent(u,~x1),\\ eats(x1,~x2,~x3)}{
$\lnot$needtoexpr(pred-eats,~x1,~x2,~x3), \\
 subst(t-28-eats,~n1,subj), subst(t28-eats,~n4,obj),\\
adj(t-28-eats, n2, u), adj(t-28-eats, n3, u)}{8.5172}\\\smallskip

@@

  A straightforward way to deal with data sparseness is to drop all lexicalizations from event definitions, as illustrated in figure \ref{fig:modelillustration}, A. Unfortunately this model does no longer account for bilexical dependencies between words. Since our system has to add a lexicalized tree in each step, lexicalizations for this child tree should always be taken into account by the probability model, if available. Despite these drawbacks, we perform experiments with the unlexicalized model as a baseline. This allows us to investigate if purely syntactic information is sufficient to achieve high quality generation output. \\  
An alternative model computes linear interpolation between three back-off levels. The first level is just standard {\sc ptag} (figure \ref{fig:modelillustration}, B.1), for the second level the lexicalizations of the parent tree is dropped (figure \ref{fig:modelillustration}, B.2), for the third level the model describes only the distribution of lexicalized child trees over each category (figure \ref{fig:modelillustration}, B.3).  
Notice also, that the third level is similar in spirit to a probabilistic version of original {\sc crisp} operators (compare figure \ref{fig:modelillustration}, B.3 and figure \ref{fig:crisp-operators}. 
\begin{figure}[t]
\begin{center}
\includegraphics[width=.5\textwidth]{figures/modelillustration}
\caption{\label{fig:modelillustration} Illustration of the unlexicalized probability model (A) and the three back-off levels of the linear interpolation model (B). B.1 is the original {\sc ptag} definition and corresponds directly to {\sc pcrisp} operators as in figure \ref{example-action}, while B.3 resembles {\sc crisp} operators (c.f. figure \ref{fig:crisp-operators}).}
